Plant Metagenomic Barcoding using Angiosperms353 of Corbiculae from wild Bumble Bees

### 2023-05-23 script/manuscript_MEE.Rmd
mit
adding JBF ST intro edits

--- 
title: Plant Metagenomic Barcoding using Angiosperms353 of Corbiculae from wild Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$, Emily J. Woodworth$^1$$^,$$^2$, 
abstract:  |  
  1) DNA Barcoding has been successful for the rapid analysis of complex ecological assemblages. However, barcoding in the plant kingdom has been more difficult than in others, limiting the applications of eDNA and metagenomics in ecology. 
  2) Here we test the use of Angiosperms 353 probes to barcode plant species found in corbiculae pollen loads collected from wild foraging bumble bees. 
  3) Using a high resolution long term observation study of wild bumble bees, we develop a simple framework for the application of metagenomics in plant-pollinator interactions to increase the efficiency and accuracy of applications using species distribution models, phenological modelling, and simple procedural decision making. 
  4) By utilizing Species distribution modelling we allow users to create regionally appropriate sequence databases which are poised to increase accuracy of the sequence assignments and  which minimize the need for large computational power, and run time. 
  5) We show that the Angiosperms 353 probes, developed for phylogenomics, and which are currently being used in the largest ever plant systematic endeavor, offer significant promise to metagenomic approaches around the globe. 
  6) The DNA barcoding of bumble bee corbiculae pollen loads was most accurate when combined with knowledge of what plant species were flowering in the plant community when they were collected. Thus, supplementing DNA barcoding data with ecological context is most accurate and powerful.
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

An enormous amount of Earths biodiversity may be attributed to the interactions between species (@soltis2019darwin, @futuyma2009macroevolution, @voje2015role, @weber2017evolution, @hembry2020ecological). These interactions not only lead to the the origin of many species but appear essential to the maintenance of virtually all ecosystems (@agrawal2007filling, @valiente2015beyond, @bascompte2006asymmetric). In order to understand and rationally conserve both the species and the ecosystems which their interactions compose - Darwins 'Entangled Bank'- integrative approaches, with the potential for scaling are required immediately (@darwin2004origin, @thompson1994coevolutionary, @agrawal2007filling, @banerjee2022environmental, @blanchet2020co, @jordano2016sampling). Major limitations, imposed by a lack of taxonomic expertise, impede our initial ability to identify organisms let alone more complex phenomena (@hebert2003biological). The lack of an ability to identify whole organisms, especially those from diverse clades where species are oftentimes delineated along ecological lines - or on occasion cryptic, lessens the utility of them to serve as bioindicators; a role they are typically pre-disposed to (@Gage2013HistoricalRO, @banerjee2022environmental, @janzen2017nuclear, @oliver2009cryptic). The ability to identify fragments of organisms (e.g. leaf tissue) increases our ability to understand the interactions of  not only entire ecosystems but also a focal, generally rare and hence difficult to detect, organism with their surrounding; allowing for the most precise allocation of conservation decisions and funds e.g. those for restoration processes (@banerjee2022environmental, @johnson2023environmental).

  Recently barcoding (the identification of a sample from a single organism *e.g.* a piece of leaf), and metabarcoding (the identification of a sample containing a mix of organisms *e.g.* soil), have shown considerable promise in all Kingdoms of Life (@ruppert2019past). With plants the identification of members of certain clades using barcoding has been quite successful (@kress2017plant), whereas many other clades have proven more problematic (@liu2014identification, @china2011comparative, @coissac2012bioinformatic), however metabarcoding incurs additional challenges to those which  exist for the currently implemented barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular challenges with the high copy number barcodes (e.g. ITS2, *rbcL*, *matK*, *trnH-psbA*) include the utilization their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant). Currently, most plant metabarcoding endeavors only allow the identification of material to the level of family or genus.

@@

  As species interactions vary both in space and time contrasts in the flowering periods of many plant species, can provide an additional filter for increasing the accuracy of identifying material in certain types of metagenomic samples (@janzen1967synchronization, @Newstrom1994ANC, @thompson1994coevolutionary, @caradonna2021seeing). In high elevation temperate regions, pollination interactions vary temporally and are characterized by high turnover in active periods of species (@caradonna2017interaction, @caradonna2020temporal), however the overall shorter extent of the active growing season in these systems results in the presence of few to any natural breaks, which reduces the utility of these to operate as filters in the post-processing of sequence matches. Nonetheless, we work develop a general approach which seems applicable to many areas which utilize the temporal dimension for classifying sequences in metagenomic samples (but see @davis2022new, Figure 1).

@@

  We tested the proposed metagenomic and informatics approach to examine the diets of wild bumblebee pollinators, and compare this approach with direct observations and the pollen record, which has shown some incongruency in several floral visitation networks involving smaller bodied fauna (@olesen2011missing, @barker2021pollen, @zhao2019topology, @alarcon2010congruence). The assessment of the plant species in pollen is a commonly desired, across several potential applications, despite numerous potential complications (@pornon2017dna, @bell2017applying, @sickel2015increased, @bell2019quantitative, @suchan2019pollen,  @johnson2021airborne). The two foraging phases of the queen bumblebee life cycle are essential to 1) increase their weight before diapause, and 2) increase their ovary weights while establishing their recently found nests. Both of these time periods represent potential demographic bottlenecks in bumble bee populations (@sarro2022bumble). Bumblebees are one of the only groups of insects with unequivocal quantitative evidence for numerous population declines, while simultaneously serving as the most effective pollinators in many temperate montane ecosystems (@cameron2020global, @goulson2008decline, @williams1982distribution, @colla2012assessing, @bergman1996micrometeorological, @bingham1998efficient, @grixti2009decline). These areas often represent the most diverse areas in the temperate and often offer the sole potential refugia for multiple dimensions of biodiversity under climate change, whilst simultaneously experiencing the greatest proportional changes in mean annual temperature (@brito2018climate, @pepin2022climate). An immediate understanding of how to manage previously overlooked keystone insect species, such as bumble bees, is essential if the refugial potential of the temperate mountains are to be incorporated into their current diversity (@loarie2009velocity, @dobrowski2016climate).

@@

To qualitatively identify, and quantitatively note, the plant species present in corbiculae loads microscopy was used. A pollen reference library of fuchsin-jelly stained grains which may be present in corbiculae loads of slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@beattie1971technique, @brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (@ogilvie2022shifting), as well as the RMBL Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non-accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (Appendix 6). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner.

@@

In 2015 the six sites were surveyed once weekly from `r s`-`r e` for a total of `r toString(sum(obs_time$tot.obs.length)/60)` hours. A total of `r toString(sum(bee_obs$interactions))` overwinter queen-pollen foraging interactions were observed (range per bee species by week range = `r toString(min(t_b_wk$n))`-`r toString(max(t_b_wk$n))`, $\bar{x}$  = `r toString(round(mean(t_b_wk$n), 2))`, median = `r toString(median(t_b_wk$n))`), with a range of total observed interactions per bee species across this time period (minimum = `r toString(min(t_b_b$n))`, $\bar{x}$  = `r toString(round(mean(t_b_b$n), 2))`, median = `r toString(median(t_b_b$n))`, max = `r toString(max(t_b_b$n))`). Plants varied widely in the number of interactions which they partook in with each species of bee (range per plant species by week minimum = `r toString(min(t_b_wk_pl$n))` - `r toString(max(t_b_wk_pl$n))`, $\bar{x}$  = `r toString(round(mean(t_b_wk_pl$n), 2))`, median = `r toString(median(t_b_wk_pl$n))`), with a range of total observed interactions per plant species over this time period (minimum = `r toString(min(t_b_pl$n))`, $\bar{x}$  = `r toString(round(mean(t_b_pl$n), 2))`, median = `r toString(median(t_b_pl$n))`, max = `r toString(max(t_b_pl$n))`). The number of plant species which bees were observed interacting with varied more narrowly (range = `r toString(min(t_spp_i$n))` - `r toString(max(t_spp_i$n))`, $\bar{x}$  = `r toString(round(mean(t_spp_i$n), 2))`, median = `r toString(median(t_spp_i$n))`), interactions were observed with a total of `r no_plant_sp` plant species.

@@

## 3.1 | Spatial Analyses

@@

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae less Heliantheae. Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`; however, Asteraceae `r toString(ff)`. Both morphotypes of Asteraceae pollen `r toString(dd)`.

@@

To ascertain the extent to which records of multiple species in a family, which were suspected to be sampling artefacts occurred in molecular samples an index of similarity, ala Jaccard, the affinity index was used to assess co-occurrence (@mainali2022better, @mainali2022alpha). Numerous taxa from the family Ranunculaceae Jussieu (*Caltha* L. sp., *Thalictrum* L. spp., *Trollius* L. sp., *Aquilegia* L. spp.), had $\alpha$ scores which indicated that they are only present when a more common confamilial taxa *Delphinium barbeyi* (Huth) Huth *nuttallianum* Pritz. were recorded. A similar relationship was observed in the Hydrophyllaceae R.Br. with samples placed in *Nemophila* Nutt., which only occurred when the more abundant *Hydrophyllum* L. species were present. The size of flower of *Nemophila breviflora* A. Gray make it unlikely to be visited by Bumble Bees, and it is a false positive. The floral morphology and orientation of flower of *Thalictrum*  spp. also makes them unlikely to be visited, and while evidence of visits to *Caltha* and *Trollius* are lacking, due to the association between the reads these results appear unlikely.

@@

Also regarding the case study, some foraging preferences of *Bombus*, at this field site and across several localities globally emerge. These suggest the need for land managers to maintain relatively high amounts of members of the plant families Fabaceae, Boraginaceae, and Ranunculaceae, in Western North American montane landscapes (@goulson2005causes, @goulson2010bumblebees, @liang2021evolutionary, @bontvsutvsnaja2021bumble). Numerous historic, and some ongoing, practices reduce the ability of many landscapes to support stable populations of *Bombus*. Historic livestock grazing was often associated with the targeted removal of many species known to have compounds toxic to cattle. In particular, the removal of locoweeds (Fabaceae: *Astragalus* L. & *Oxytropis* DC.) and larkspurs (Ranunculaceae: *Delphinium*) were common across public lands administered by the U.S. Forest Service (@ralphs1988herbicide, @aldous1919eradicating, @ralphs2003mechanism). Further actions, generally initiated by early settlers, involved the channelization and incising of streams, culling of beavers, and leaving cattle concentrated on higher order stream banks, processes which lowered water tables and reduced the extent of stream-associated wetlands and the mesic meadows fringes which provide habitat for many species of 'tall' *Mertensia* (Boraginaceae, e.g. *M. ciliata* Torr. G. Don.), to an extent *Delphinium barbeyi*  and many species of native *Trifolium* L. (@dahl1990wetlands, @naiman1988alteration, @belsky1999survey, @cooke1976arroyos). Fire suppression resulted in the succession of many Aspen (*Populus tremuloides* Michx.) groves to Conifer stands, decreasing the mosaic of age structured habitats in many landscapes, adversely effecting habitat for tall *Mertensia* species and several species of *Delphinium* (@brewen202176, @keane2002cascading). Finally the effects of Nitrogen deposition, given the West's rapidly growing population still pose adverse effects on the abundance of species of Fabaceae at urban-rural interfaces (see @stevens2018atmospheric, @fenn2003ecological). Current solutions to the above issues, involve targeted burns, reintroduction of beavers and beaver habitat analogs, and the possibility of re-seeding a variety of 'locoweeds' and 'larkspurs' in areas now seldom used, or only used for early, grazing. The highly enthusiastic response of land managers, and homeowners, to plant *Ascelpias* L., using genetically appropriate materials, to improve Monarch Butterfly (*Danaus plexippus* L.) habitat provides an effective framework for the latter (@oberhauser2015monarchs, @basey2015producing).

@@

We believe that the combination of spatial and temporal models, united and guided by localized natural history knowledge, provides the essential components of a bayesian framework for approaching the coarse elucidation of ecological interactions using DNA Barcoding. Herein we crudely utilized this thinking via binary outcomes, should a species predicted be predicted present or not? Is it unequivocally flowering or not? Myriad data show biological systems and ecological interactions have more variance than can be reasonably discretely parsed. We expect that within a bayesian framework studies of pollinator behavior may be enacted via this approach at a landscape level, e.g. the scale of an entire drainage basin such as the Gunnison which is quickly becoming one of the worlds few model ecosystems. We hope that the A353 probes as tools for metabarcoding play a role in these endeavors.

>>>>

--- 
title: Plant Metagenomic Barcoding using Angiosperms353 of Corbiculae from wild Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$, Emily J. Woodworth$^1$$^,$$^2$, 
abstract:  |  
  1) DNA Barcoding has been successful for the rapid identification of species in complex ecological assemblages. However, metabarcoding in the plant kingdom has been limited due to a lack of universal gene regions that works across all taxa, limiting the applications of eDNA and metagenomics in ecology.  
  2) To circumvent these limitations, we propose using more holistic approach which combines a multiple gene approach that incorporates existing plant occurrence databases, species distribution models, and phenology analyses to generate a list of likely candidate species to increase accuracy. Such an effort can help guide plant sample collection and library construction, particularly for projects with limited resources or local taxonomic knowledge. In addition, we demonstrate that building a custom candidate plant list based on known occurrences, ecological requirements, and phenology can improve computing efficiency, speed, and accuracy of metabarcoding. We tested this approach to identify the plant species contributing to the corbiculae pollen loads of bumble bees. To validate the efficiency of this approach we compare results of the DNA barcoding to expert-led pollen identification and field observations of plant-pollinator interactions that generated these pollen loads.   
  3) We show that the Angiosperms 353 probes, developed for phylogenomics, and which are currently being used in the largest ever plant systematic endeavor, offer significant promise to metagenomic approaches around the globe. The DNA barcoding of bumble bee corbiculae pollen loads was most accurate when combined with knowledge of what plant species were flowering in the plant community when they were collected.   
  4) Supplementing DNA barcoding data with ecological context is most accurate and powerful.  
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

Large scale species loss (@joppa2011biodiversity; @pimm2014biodiversity) and biotic homogeneization (REF), and the impacts of these processes on ecosystem functions and human well-being (@cardinale2012biodiversity), have inspired numerous calls for a more consistent monitoring of ecosystems and their diversity (@pereira2006towards; @pecl2017biodiversity). Monitoring ecological communities structure and interaction will be critical to informing and prioritizing conservation efforts (REF), particularly considering the large proportion of know species currently threaten or likely to be threaten in the near future (@pimm2014biodiversity; @pereira2010scenarios). However, ecosystem monitoring remains inconsistent in time and space (@yoccoz2001monitoring) as it requires substantial resources, thereby leaving several regions, ecosystems, and even species under-observed (@pereira2012global; @collen2008tropical; @meyer2015global; @ruete2015displaying).

Species interactions are a critical component of ecosystem stability and therefore biodiversity conservation (Soltis et al. (2019), @futuyma2009macroevolution, Voje et al. (2015), Weber et al. (2017), @hembry2020ecological). Species interactions, which span from positive (e.g., symbiosis, mutualistic) to negative (e.g., predation, competition), shape species distribution (@bascompte2009mutualistic; Wisz et al. 2013) and tolerance to environmental stressors (REF). They impact ecosystem stability and diversity (@agrawal2007filling, @valiente2015beyond, @bascompte2006asymmetric, and the relationships between biodiversity and ecosystem functions (@duffy2007functional). However, species interactions are challenging to observe (REFs). They are typically measured through shorter-term observational studies yet are highly variable in time and space (@liu2022environmental). Studying species interactions also requires extensive taxonomic expertise, especially those from diverse clades (@hebert2003biological), further challenging their frequent assessment. As a result, few ecological networks have been fully mapped (REFs). Improving our capacity to efficiently study species interactions consistently across several ecosystems, landscapes, and management context is thus pivotal to enacting conservation interventions (@tylianakis2010conservation). This is particularly urgent as increased biological introductions and climate change are impacting species interactions at a rapid rate, thus calling for more frequent and consistent monitoring (REFs).

  Recently barcoding (the identification of a sample from a single organism *e.g.* a piece of leaf), and metabarcoding (the identification of a sample containing a mix of organisms *e.g.* soil), have shown considerable promise in all Kingdoms of Life (@ruppert2019past). The ability to identify species from a fragments of organisms (e.g. hair, scat, leaf tissue) increases our ability to understand the interactions of not only entire ecosystems but also a focal, generally rare and hence difficult to detect, organism with their surrounding; allowing for the most precise allocation of conservation decisions and funds e.g. those for restoration processes (Banerjee et al. (2022), Johnson et al. (2023)). With plants the identification of members of certain clades to species using barcoding has been successful in some groups (@kress2017plant), whereas many other clades have proven more problematic (@liu2014identification, @china2011comparative, @coissac2012bioinformatic). Metabarcoding incurs additional challenges to those which exist for the currently implemented barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular limitations of the traditional high copy number barcodes (e.g. ITS2, *rbcL*, *matK*, *trnH-psbA*) include the utilization their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant). Currently, most plant metabarcoding endeavors only allow the identification of material to the level of family or genus.

@@

  Plant flowering varies over space and also in time, and in particular, contrasts in the flowering periods of many plant species can provide an additional filter for identifying material in certain types of metagenomic samples (@janzen1967synchronization, @Newstrom1994ANC, @thompson1994coevolutionary, @caradonna2021seeing). In high elevation temperate regions, pollination interactions vary temporally and are characterized by high turnover in active periods of species (@caradonna2017interaction, @caradonna2020temporal). We work to develop a general approach which seems applicable to many areas which utilize the temporal dimension for classifying sequences in metagenomic samples (but see @davis2022new, Figure 1).

@@

  We tested this metagenomic and informatics approach by examining the plant species composition within corbiculae pollen loads of bumble bees as a case study. In particular, we use DNA barcoding to identify the plant species in corbiculae pollen loads collected from overwintered queen bumble bees, and compare this approach with direct observations of bee flower visits and microscopically-examined pollen slides, which has shown some incongruency in several floral visitation networks involving smaller bodied fauna (@olesen2011missing, @barker2021pollen, @zhao2019topology, @alarcon2010congruence). The assessment of the plant species in pollen is a commonly desired, across several potential applications, despite numerous potential complications (@pornon2017dna, @bell2017applying, @sickel2015increased, @bell2019quantitative, @suchan2019pollen,  @johnson2021airborne). The two foraging phases of the queen bumblebee life cycle are essential to 1) increase their weight before diapause, and 2) increase their ovary weights while establishing their recently found nests. Both time periods represent potential demographic bottlenecks in bumble bee populations (@sarro2022bumble). Bumblebees are one of the only groups of insects with unequivocal quantitative evidence for numerous population declines, while simultaneously serving as the most effective pollinators in many temperate montane ecosystems (@cameron2020global, @goulson2008decline, @williams1982distribution, @colla2012assessing, @bergman1996micrometeorological, @bingham1998efficient, @grixti2009decline). These areas often represent the most diverse areas in the temperate and often offer the sole potential refugia for multiple dimensions of biodiversity under climate change, whilst simultaneously experiencing the greatest proportional changes in mean annual temperature (@brito2018climate, @pepin2022climate). An immediate understanding of how to manage previously overlooked keystone insect species, such as bumble bees, is essential if the refugial potential of the temperate mountains are to be incorporated into their current diversity (@loarie2009velocity, @dobrowski2016climate).

@@

To qualitatively identify, and quantitatively note, the plant species present in corbiculae loads microscopy was used. A pollen reference library of fuchsin-jelly stained grains which may be present in corbiculae loads of slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@beattie1971technique, @brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (@ogilvie2022shifting), as well as the RMBL Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non-accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used clustering techniques to supplement our subjective opinions of which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (Appendix 6). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner.

@@

In 2015 the six sites were surveyed once weekly from `r s`-`r e` for a total of `r toString(sum(obs_time$tot.obs.length)/60)` hours. A total of `r toString(sum(bee_obs$interactions))` overwintered queen-pollen foraging interactions were observed (range per bee species by week range = `r toString(min(t_b_wk$n))`-`r toString(max(t_b_wk$n))`, $\bar{x}$  = `r toString(round(mean(t_b_wk$n), 2))`, median = `r toString(median(t_b_wk$n))`), with a range of total observed interactions per bee species across this time period (minimum = `r toString(min(t_b_b$n))`, $\bar{x}$  = `r toString(round(mean(t_b_b$n), 2))`, median = `r toString(median(t_b_b$n))`, max = `r toString(max(t_b_b$n))`). Plants varied widely in the number of interactions which they partook in with each species of bee (range per plant species by week minimum = `r toString(min(t_b_wk_pl$n))` - `r toString(max(t_b_wk_pl$n))`, $\bar{x}$  = `r toString(round(mean(t_b_wk_pl$n), 2))`, median = `r toString(median(t_b_wk_pl$n))`), with a range of total observed interactions per plant species over this time period (minimum = `r toString(min(t_b_pl$n))`, $\bar{x}$  = `r toString(round(mean(t_b_pl$n), 2))`, median = `r toString(median(t_b_pl$n))`, max = `r toString(max(t_b_pl$n))`). The number of plant species which bees were observed interacting with varied more narrowly (range = `r toString(min(t_spp_i$n))` - `r toString(max(t_spp_i$n))`, $\bar{x}$  = `r toString(round(mean(t_spp_i$n), 2))`, median = `r toString(median(t_spp_i$n))`), interactions were observed with a total of `r no_plant_sp` plant species.

@@

## 3.2 | Spatial Analyses

@@

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae less the Heliantheae Aliiance (Sunflower Tribe) (Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`; however, Asteraceae `r toString(ff)`. Both morphotypes of Asteraceae pollen `r toString(dd)`.

@@

A number of suspected false positives were returned by the sequence classified, in all instances these were species present in low abundances at the field station which were also returned with a common species in the same family. To ascertain the extent to which records of multiple species in a family, which were suspected to be sampling artefacts occurred in molecular samples an index of similarity, ala Jaccard, the affinity index was used to assess co-occurrence (@mainali2022better, @mainali2022alpha). Numerous taxa from the family Ranunculaceae Jussieu (*Caltha* L. sp., *Thalictrum* L. spp., *Trollius* L. sp., *Aquilegia* L. spp.), had $\alpha$ scores which indicated that they are only present when a more common confamilial taxa *Delphinium barbeyi* (Huth) Huth *nuttallianum* Pritz. were recorded. A similar relationship was observed in the Hydrophyllaceae R.Br. with samples placed in *Nemophila* Nutt., which only occurred when the more abundant *Hydrophyllum* L. species were present. The flower size  and sparsity of *Nemophila breviflora* A. Gray make it unlikely to be visited by Bumble Bees, and it is a false positive. The floral morphology and orientation of flower of *Thalictrum*  spp. also makes them unlikely to be visited, and while evidence of visits to *Caltha* and *Trollius* are lacking, due to the association between the reads these results appear unlikely.

@@

Also regarding the case study, some foraging preferences of *Bombus*, at this field site and across several localities globally emerge. These suggest the need for land managers to maintain relatively high amounts of members of the plant families Fabaceae, Boraginaceae, and Ranunculaceae, in Western North American montane landscapes (@goulson2005causes, @goulson2010bumblebees, @liang2021evolutionary, @bontvsutvsnaja2021bumble). Numerous historic  practices reduce the ability of many landscapes to support stable populations of *Bombus*. Historic livestock grazing involved the removal of many species known to have compounds toxic to cattle. In particular, the removal of locoweeds (*Astragalus* & *Oxytropis*) and larkspurs (*Delphinium*) were common across public lands administered by the U.S. Forest Service (@ralphs1988herbicide, @aldous1919eradicating, @ralphs2003mechanism). Further actions, generally initiated by early settlers, involved the channelization and incising of streams, culling of beavers, and leaving cattle concentrated on higher order stream banks, processes which lowered water tables and reduced the extent of stream-associated wetlands and the mesic meadows fringes which provide habitat for many species of 'tall' *Mertensia* (e.g. *M. ciliata*), to an extent *Delphinium barbeyi*  and many species of native *Trifolium* (@dahl1990wetlands, @naiman1988alteration, @belsky1999survey, @cooke1976arroyos). Fire suppression resulted in the succession of many Aspen (*Populus tremuloides*) groves to Conifer stands, decreasing the mosaic of age structured habitats in many landscapes, adversely effecting habitat for tall *Mertensia* species and several species of *Delphinium* (@brewen202176, @keane2002cascading). Finally the effects of Nitrogen deposition, given the West's rapidly growing population still pose adverse effects on the abundance of species of Fabaceae at urban-rural interfaces (see @stevens2018atmospheric, @fenn2003ecological). Current solutions to the above issues, involve targeted burns, reintroduction of beavers and beaver habitat analogs, and the possibility of re-seeding a variety of 'locoweeds' and 'larkspurs' in areas now seldom used, or only used for early, grazing. The highly enthusiastic response of land managers, and homeowners, to plant *Ascelpias*, using genetically appropriate materials, to improve Monarch Butterfly (*Danaus plexippus*) habitat provides an effective framework for the latter (@oberhauser2015monarchs, @basey2015producing).

@@

We believe that the combination of spatial and temporal models, united and guided by localized natural history knowledge, provides the essential components of a framework for approaching the coarse elucidation of ecological interactions using DNA Barcoding. Herein we crudely utilized this thinking via binary outcomes, should a species predicted be predicted present or not? Is it unequivocally flowering or not? Myriad data show biological systems and ecological interactions have more variance than can be reasonably discretely parsed. We expect that within a framework developed from our preliminary works studies of pollinator behavior may be enacted via this approach at a landscape level, e.g. the scale of an entire drainage basin such as the Gunnison which is quickly becoming one of the worlds few model ecosystems. We hope that the A353 probes as tools for metabarcoding play a role in these endeavors.
